A Continual Planner that Switches Between Serial and Decision-Theoretic Planning

### 2010-11-17 subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/moritz_2011.tex
I'll try and work on this again tonight. I'm not happy with how I think the paper is going to end up, but I don't know what else to do. I think we need to make the introduction and discussion sections as strong as possible. Just make it an interesting discussion of the literature here.
And make sure the experiments show our sensing is better!!

 - Charles

We describe the partially observable propositional probabilistic
planning problem, with costs and rewards. The underlying world
dynamics are modelled in terms of a finite set of stochastic actions
$\stochActions$, deterministic outcomes $\detActions$, and
state-characterising propositions $\propositions$.

A problem state $\state$ is a set of propositions $\state \subseteq
\props$. Every stochastic action $\stochAction \in \stochActions$ has
a precondition $\poss(\stochAction)$, also a set of propositions. An
action is applicable at a state $\state$ when $\poss(\stochAction)
\subseteq \state$. We denote $\stochActions(\state)$ the set of
actions applicable at state $\state$.  When $\stochAction \in
\stochActions(\state)$ is applied, nature chooses one element amongst
a small set of deterministic outcomes $\detActions(\stochAction)
\equiv \{\detAction_1, \ldots, \detAction_k \}$. We denote
$\mu_{\stochAction}(\detAction_i)$ the probability that nature takes
outcome $\detAct_i$, and for all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$. That outcome so chosen has an
effect on the underlying state given in terms of two lists of
propositions. The add-list $\add(\detAction)$ and delete-list
$\delete(\detAction)$.\footnote{If a proposition is in the add-list of
  $\detAction$, then it cannot be in the delete-list and vice versa.}
If outcome $\detAction$ with $\add(\detAction) := [\prop_1,
  ..,\prop_n]$ and $\delete(\detAction) := [\prop_1, ..,\prop_m]$ is
chosen by nature when \stochAction\ is applied at state $\state$, then
the resultant state is $ ( \state \cup \add(\detAction) ) \backslash
\delete(\detAction)$ -- i.e., propositions from $\add(\detAction)$ are
added to $\state$, and those from $\delete(\detAction)$ are removed
from $\state$.

We are concerned with problems that feature partial
observability. These have a perceptual model given in terms of a
finite set of stochastic senses $\stochSenses$, deterministic sensing
outcomes $\detSenses$, and perceptual propositions $\percepts$, called
percepts. Here, an observation $\observ$ is a set of percepts $\observ
\subseteq \percepts$, and we denote \observations\ the set of
reachable observations. The underlying state of the world cannot be
observed directly, rather, senses $\stochSense \in \stochSenses$
effect an observation $\observ \in \observations$ that informs what
should be believed about the world. In the POMDP model, if $\action$
is applied effecting a transition to a successor state $\state'$, then
an observation occurs according to the active senses
$\stochSenses(\stochAction, \state') \subseteq \stochSenses$. A
precept is active, written $\stochSense \in \stochSenses(\stochAction,
\state')$, if the senses action-precondition,
$\poss_\stochActions(\stochSense)$, is equal to $\stochAction$, and
the state-precondition $\poss_\states(\stochSense) \subseteq
\propositions$ is satisfied by the state $\state'$; In the usual sense
that $\poss_\states(\stochSense) \subseteq \state'$.
%%
When a sense is active, nature must choose exactly one outcome amongst
a small set of deterministic choices $\detSenses(\stochSense)
\equiv \{\detSense_1, \ldots, \detSense_k \}$, so that for each
$i$ we have $\detSense_i \subseteq \percepts$. The probability of
the $i^{th}$ element being chosen is given by
$\psi_{\stochSense}(\detSense_i)$, where $\sum_{\detSense_i \in
\detSenses(\stochSense)} \psi_{\stochSense}(\detSense_i) =
1$. The observation received by the agent corresponds to the union of
perceptual propositions from chosen elements of active senses.

A POMDP has a starting configuration that corresponds to a subjective
Bayesian belief state. Written $\bstate_0$, this is a probability
distribution over states. We use the notation $\bstate_0(\state)$ to
denote the probability that the starting state is equal to
$\state$. We suppose this is given in a factored tree format.

%% Let $\detSense_i \subseteq \percepts$ for each $i$, so that if
%% we apply $\stochAction$ and transition to $\state'$, the observation
%% received will be a union of choices from the active percepts, which
%% will occur with the probability according to $\psi$ that nature can
%% make those choices.

Finally, we make a number of fairly standard assumptions. First, that
action execution and sensing occurs instantaneously, and that only one
action can be applied at a plan-step. Second, it can arise that in
some propositional states an action is applicable, and in others it is
not. Moreover, a belief-state can assign non-zero probability to
states where that applicability holds, and states where it does
not. We differ
from~\citeauthor{younes:littman:04}~\citeyear{younes:littman:04},
because we forbid the execution of actions that are not applicable in
any state of the current belief.  Also,
unlike~\citeauthor{hoffmann:brafman:2006}~\citeyear{hoffmann:brafman:2006},
we do incorporate a PPDDL-like default semantics, supposing that when
an action \stochAction\ is executed at some \bstate, that it has no
effect on states $\bstate(\state) > 0$ where $\poss(\stochAction)
\not\subseteq \state$.

@@

Using the notation \bstate\ for a belief-state, and writing
$\bstate(\state)$ for the probability the robot believes it is in
state \state, we have that the value of that belief $\bstate$ according to a
controller is:

@@

The modelling language of choice for planning in probabilistic
problems is the Probabilistic Planning Domain Definition
Language~\cite{younes:littman:04}. PPDDL was used in all 3 of the
International Planning Competitions since 2004. A variation on PDDL
for describing domains with stochastic actions and uncertain starting
configurations, PPDDL is a declarative first-order (a.k.a.,
relational) that facilitates factored description of domains and
problem instances. There are straightforward compilations from
problems expressed compactly in PPDDL to propositional representations
amenable to state-of-the-art planning procedures.  Because this
language cannot model domains that feature partial observability, for
our setting we have developed DTPDDL, an extension of PPDDL to partial
observability.

For example, $(:\pp{predicates} (\pp{location} ?p - VisualObject))$
says that there is a unary predicate

@@

\small{
\begin{tabtt}
(\= :action look-at-object \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - label ?p - place) \\
  \> :precondition (and (= (is-in ?r) ?p) \\
  \> \> (= (label ?o) ?l) ) \\
  \> :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
}

\small{
\begin{tabtt}
(\= :observe visual-object \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - label ?p - place) \\
  \> :execution (look-at-object ?r ?v ?l ?p) \\
  \> :effect (when (= (is-in ?o) ?p) (probabilistic 0.8)) \\
\end{tabtt}
}

\small{
\begin{tabtt}
(\= :sensor look-for-object \\
  \> :agent (?a - robot) \\
  \> :parameters (?l - label ?p - place) \\
  \> :variables (?o - visual-object) \\
  \> :precondition (\=and (= (is-in ?r) ?p) \\
  \> \> (= (label ?o) ?l) (assume (is-in ?o) ?p) ) \\
  \> :effect () \\
  \> :sense (= (is-in ?o) ?p) \\
\end{tabtt}
}

>>>>

We describe the partially observable propositional probabilistic
planning problem, with costs and rewards. The underlying process
dynamics are modelled in terms of a finite set of stochastic actions
$\stochActions$, deterministic outcomes $\detActions$, and
state-characterising propositions $\propositions$.

We model a process state $\state$ as the set of propositions that are
true of the state. Notationally, we have $\state \subseteq
\props$. State change is induced by application of actions. A
stochastic action $\stochAction \in \stochActions$ is applicable if
its precondition $\poss(\stochAction)$, a set of propositions, are
satisfied in the current state -- I.e., $\poss(\stochAction) \subseteq
\state$. We write $\stochActions(\state)$ for set of actions
applicable at state $\state$.  If action $\stochAction \in
\stochActions(\state)$ is applied at \state, nature chooses one
element amongst a small set of deterministic outcomes
$\detActions(\stochAction) \equiv \{\detAction_1, \ldots, \detAction_k
\}$. We denote $\mu_{\stochAction}(\detAction_i)$ the probability that
nature takes outcome $\detAct_i$, and for all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$. The chosen outcome has an
effect on the state given in terms of two lists of propositions. The
add-list $\add(\detAction)$ and delete-list
$\delete(\detAction)$.\footnote{If a proposition is in the add-list of
$\detAction$, then it cannot be in the delete-list and vice versa.}
If outcome $\detAction$ with $\add(\detAction) := [\prop_1,
..,\prop_n]$ and $\delete(\detAction) := [\prop_1, ..,\prop_m]$ is
chosen by nature when \stochAction\ is applied at state $\state$, then
the resultant state is $ ( \state \cup \add(\detAction) ) \backslash
\delete(\detAction)$ -- i.e., propositions from $\add(\detAction)$ are
added to $\state$, and those from $\delete(\detAction)$ are removed
from $\state$.

We are concerned with problems that feature partial
observability. These have a perceptual model given in terms of a
finite set of stochastic {\em senses} $\stochSenses$, deterministic
sensing outcomes $\detSenses$, and perceptual propositions
$\percepts$, called {\em percepts}. Here, an observation $\observ$ is
a set of percepts $\observ \subseteq \percepts$, and we denote
\observations\ the set of reachable observations. The underlying state
of the process cannot be observed directly, rather, senses
$\stochSense \in \stochSenses$ effect an observation $\observ \in
\observations$ that informs what should be believed about the state a
process is in. In detail, if $\action$ is applied effecting a
transition to a successor state $\state'$, then an observation occurs
according to the active senses $\stochSenses(\stochAction, \state')
\subseteq \stochSenses$. A percept is active, written $\stochSense \in
\stochSenses(\stochAction, \state')$, if the senses'
action-precondition, $\poss_\stochActions(\stochSense)$, is equal to
$\stochAction$, and the state-precondition $\poss_\states(\stochSense)
\subseteq \propositions$ is satisfied by the state $\state'$; In the
usual sense that $\poss_\states(\stochSense) \subseteq \state'$.
%%
When a sense is active, nature must choose exactly one outcome amongst
a small set of deterministic choices $\detSenses(\stochSense)
\equiv \{\detSense_1, \ldots, \detSense_k \}$, so that for each
$i$ we have $\detSense_i \subseteq \percepts$. The probability of
the $i^{th}$ element being chosen is given by
$\psi_{\stochSense}(\detSense_i)$, where $\sum_{\detSense_i \in
\detSenses(\stochSense)} \psi_{\stochSense}(\detSense_i) =
1$. The observation received by the agent corresponds to the union of
perceptual propositions from chosen elements of active senses.

A POMDP has a starting configuration that corresponds to a Bayesian
belief-state. Intuitively, this is the robot's subjective belief about
its environment. Formally, a belief-state $\bstate$ is a probability
distribution over process states. We write $\bstate(\state)$ to denote
the probability that the process is in $\state$ according to
$\bstate$, and $\bstate_0$ when discussing the starting
configuration.

Finally, we make a number of fairly standard assumptions. First, that
action execution and sensing occurs instantaneously, and that only one
action can be applied at a plan-step. Second, it can arise that in
some propositional states an action is applicable, and in others it is
not. Moreover, a belief-state can assign non-zero probability to
states where that applicability holds, and states where it does
not. We differ
from~\citeauthor{younes:littman:04}~(\citeyear{younes:littman:04}),
because we forbid the execution of actions that are not applicable in
any state of the current belief.  Also,
unlike~\citeauthor{hoffmann:brafman:2006}~(\citeyear{hoffmann:brafman:2006}),
we do incorporate a PPDDL-like default semantics, supposing that when
an action \stochAction\ is executed at $\state_i$ where
$\bstate(\state_i) > 0$, that it has no effect on states $\state_i$ if
$\poss(\stochAction) \not\subseteq \state_i$ and $\bstate(\state_i) >
0$.

@@

\noindent The value of $\bstate$ according to a controller is
therefore:

@@

The modelling language of choice for planning in probabilistic
problems is the Probabilistic Planning Domain Definition
Language~\cite{younes:littman:04,younes:etal:2005}. PPDDL was used in
all 3 of the International Planning Competitions since 2004. A
variation on PDDL for describing domains with stochastic actions and
uncertain starting configurations, PPDDL is a declarative first-order
(a.k.a., relational) language that facilitates factored description of
domains and problem instances. There are straightforward compilations
from problems expressed compactly in PPDDL to propositional
representations amenable to state-of-the-art planning procedures.
Because this language cannot model domains that feature partial
observability, for our setting we have developed DTPDDL, an extension
of PPDDL to partial observability.

Our mobile robot is able to look for a visual object, for example a
box of cornflakes, at a given place. We can model that deterministic
action using the following schema:

\def\oom{$^{\tt +}$}
\def\zom{$^*$}
\def\bump{\hspace{1cm}}
\def\req#1{$^{\tt #1}$}
\def\noteme#1{}%{[[#1]]}
\def\notecoauth#1{\ }
\def\notereader#1{[[#1]]}
\def\meta{$\uparrow\uparrow$}
\def\la{\langle}
\def\ra{\rangle}

\small{
\begin{tabtt}
(\= :action look-for-object \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - location) \\
  \> :precondition (and (= (is-in ?r) ?l) ) \\
  \> :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
}

\noindent That is, the robot can always look for a visual object at
its current location. The only effect this has on the state, is that
the robot incurs a penalty, $3$, that corresponds to the {\em cost} of
performing a visual search. Instances of this operator are useful,
because their application activates an the following sense schema:

\small{
\begin{tabtt}
(\= :sense vision \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - location) \\
  \> :execution (look-for-object ?r ?v ?l) \\
  \> :effect ( and (when (= (is-in ?v) ?l) \\
  \> \bump (probabilistic 0.8 \\
  \> \bump (= (observe-is-in ?v) ?l))) \\
  \> (when (not (= (is-in ?v) ?l)) \\
  \> \bump (probabilistic 0.1 \\
  \> \bump (= (observe-is-in ?v) ?l))))) \\
\end{tabtt}
}

\noindent Here, the {\em execution} clause is a lifted description of
the sense action-precondition -- I.e,
$(\pp{look-for-object}~\pp{R2D2}~\pp{box}~\pp{desk}) =
\poss_\stochActions(\pp{vision}~\pp{R2D2}~\pp{box}~\pp{desk})$. No
other conditions guard the activation of this sense, therefore if
$(\pp{look-for-object} \pp{Dora} \pp{cornflakes} \pp{bedroom})$ is
executed, we suppose there is a $0.8$ chance of observing the
cornflakes if they are in the $\pp{bedroom}$, and otherwise a $0.1$ of
observing them.\footnote{We have simplified the schemata for
illustrative purposes. In practice, the probabilities of a particular
observation here should be parametrised by the type of visual object
the robot is searching for, and category of the location.}

\scriptsize
\begin{nopagebreak}\begin{tabtt}
<observation-def> \=::= \=(:sense <observation symbol> \\
                        \> \> :parameters (<typed list (variable)>)  \\
                        \> \> <o-def body>) \\
  <o-symbol> \> ::= <name> \\
  <o-def body> \> ::= [:precondition <GD>] \\
  \> \> [:execution <atomic action(term)> ] \\
    \> \> [:effect <o-effect>] \\
  <atomic action(t)> \> ::= (<action symbol> t\zom) \\
  <o-effect> \> ::= (and <c-o-effect>\zom) \\
  <o-effect> \> ::= <c-o-effect> \\
  <c-o-effect> \> ::= \req{:probabilistic-effects} (probabilistic <prob> <o-effect>) \\
  <c-o-effect> \> ::= <p-o-effect> \\
  <atomic o-formula(t)> \> ::= (<observation> t\zom) \\
  <p-o-effect> \> ::= <atomic o-formula(term)> \\
  <p-o-effect> \> ::= (not <atomic o-formula(term)>) \\
  <o-f-comp> \> ::= (<binary-comp> <o-f-exp> <o-f-exp>)\\
  <o-f-exp> \> ::= <number>\\
  <o-f-exp> \> ::= (- <o-f-exp>)\\
  <o-f-exp> \> ::= <o-f-head>\\
  <o-f-head> \> ::= (<o-function-symbol> <term>\zom )\\
  <o-f-head> \> ::= <o-function-symbol>\\
\end{tabtt}\end{nopagebreak}
\noteme{<structure-def> ::= <attach-def>}
\normalsize

We suppose $\bstate_0$ is given in a factored tree format of the
form:

%% \small{
%% \begin{tabtt}
%% (:init  (and \=f_{11} f_{12} ..
%%  \> (probabilistic p_{}) 
%%  \> (probabilistic ..) ..) \\
%% \end{tabtt}
%% }

%% \[(:\pp{init} (and f_{11} f_{12} .. (\pp{probabilistic} p_) ))\]

\noindent Nodes of the tree have one of three types: (1) atomic nodes
are labelled with a state proposition, e.g., $(=
(\pp{location}~\pp{cornflakes}) \pp{kitchen})$, (2) probabilistic
nodes, have the form $(r_1, C_1;..;r_n, C_n)$, and (3) constructive
nodes, take the form of a set of probabilistic and atomic nodes. The
root node is conjunctive, and leaves are always atomic. A convenient
way to give semantics to the tree, is to treat nodes as if they were
actions. An atomic node is an action with a positive effect
corresponding to the node label, .

For example, $(:\pp{predicates} (\pp{location} ?l - VisualObject))$
says that there is a unary predicate

@@

\small{
\begin{tabtt}
(\= :action look-at-object \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - label ?l - location) \\
  \> :precondition (and (= (is-in ?r) ?l) \\
  \> \> (= (label ?o) ?l) ) \\
  \> :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
}

\small{
\begin{tabtt}
(\= :observe visual-object \\
  \> :parameters \=(?r - robot ?v - visual-object\\
  \> \> ?l - label ?l - location) \\
  \> :execution (look-at-object ?r ?v ?l ?l) \\
  \> :effect (when (= (is-in ?o) ?l) (probabilistic 0.8)) \\
\end{tabtt}
}

\small{
\begin{tabtt}
(\= :sensor look-for-object \\
  \> :agent (?a - robot) \\
  \> :parameters (?l - label ?l - location) \\
  \> :variables (?o - visual-object) \\
  \> :precondition (\=and (= (is-in ?r) ?l) \\
  \> \> (= (label ?o) ?l) (assume (is-in ?o) ?l) ) \\
  \> :effect () \\
  \> :sense (= (is-in ?o) ?l) \\
\end{tabtt}
}
